---
title: "Generalized Single Step Genomic Selection Tutorial"
author: "Miguel Pérez-Enciso (mperezenciso@gmail.com)"
output: 
  learnr::tutorial:
    toc: true
    toc_depth: 3
    toc_float: true
runtime: shiny_prerendered
---

## **Outline**

We present a generalized single step (GSS) method that can accommodate any number of hierarchical molecular datasets (e.g., sequence, high and low density arrays) and pedigree, avoiding imputation. We proof that a similar efficient inversion algorithm exists. The method is recursive, starting with the highest marker density scenario.


```{r setup, include=FALSE}
library(MASS)
library(learnr)
library(AlphaSimR)
library(devtools)
library(here)

knitr::opts_chunk$set(echo = TRUE)

# This sets a seed for the entire knitting process
set.seed(42)

# auxiliary functions
source_url("https://raw.githubusercontent.com/miguelperezenciso/GSS/main/aux.R")

# ============================================================
# alphasimR wrapper
# ============================================================
# simulates nGen cycles of nGen discrete generations with 
# random crossings 
#  - h2: heritability
#  - nFounder: n founders
#  - nOffspring: per cross
#  - nChr: no chromosomes
#  - nQTL per chr
#  - nSNP per chr
# ============================================================
#------------------------------------------------------------------------
runAlphasim <- function(h2, nFounder, nOffspring, nChr, nQTL, nSNP, nGen) {
#------------------------------------------------------------------------
   #--> Founder population
   founderPop = runMacs(nInd = nFounder,  
                        nChr = nChr,   
                        segSites = nSNP)
   
   # ULL: SP is now global variable
   SP <<- SimParam$new(founderPop)
   SP$addTraitA(nQtlPerChr = nQTL)   # add QTL
   SP$setVarE(h2 = h2)               # sets error variance to achieve desired h²

   #--> mating cycles
   pops = list()
   pops[[1]] = newPop(founderPop) 
   for (t in 1:nGen) {
      # Random mating 
      pops[[t+1]] <- randCross(pops[[t]], 
                               nCrosses = nFounder,
                               nProgeny = nOffspring,
                               simParam = SP)
   }
   # merge all generations
   pop = Reduce(c, pops)

   #--> Assign phenotypes 
   pop = setPheno(pop, h2 = h2, simParam = SP)
   y = pop@pheno

   #--> Get all segregating SNPs + QTL
   g = pullSegSiteGeno(pop, simParam = SP)
   
   #--> Get all segregating SNPs + QTL
   q = pullQtlGeno(pop, simParam = SP) 
   
   #--> Get QTL genotypes
   tbv = pop@gv

   #--> get ped, add generation column
   ped = data.frame(pop@iid, as.integer(pop@father), as.integer(pop@mother))
   ped <- cbind( ped, c(rep(0,nFounder), sort(rep(1:nGen,nFounder*nOffspring))) )

   # returns pedigree, phenotypes, sequence, qtl genotypes, true breeding value
   return(list(ped=ped, y=y, g=g, q=q, tbv=tbv))
}

# ======================================================
# Computes H matrix for any number of hierarchical GRMs
# ======================================================
#------------------------
doHgss <- function(Glist) {
#------------------------
  # Input is a list of molecular relationship matrices in increasing marker density
  # Output is the covariance H matrix
  # WARNING: G list is in decreasing size order
  #          G[[1]] supposed to be NRM but could be the lowest density marker GRM
  #                 if pedigree unavailable
  # All individuals in G[i+1] are to be in G[i] and in same order
  # All markers in G[i] are supposed to be in G[i+1]
  nb = length(Glist)     # n G's
  n  = ncol(Glist[[1]])  # ninds
  H  = matrix(0, nrow=n, ncol=n)
  
  n0 = ncol(Glist[[nb]])
  # This is the span of inds for highest density genotyped
  H[(n-n0+1):n, (n-n0+1):n] = Glist[[nb]]
  for (i in (nb-1):1) {
    G  = Glist[[i]]   # G of current block
    n1 = ncol(G)      # coordinates wrt to current G
    b1 = 1:(n1-n0)    # individuals with current array ungenotyped with former denser array
    b0 = (n1-n0+1):n1 # previous individuals genotyped with denser array
    h1 = b1 + (n-n1)  # coordinates in global H: 1...nind
    h0 = b0 + (n-n1)
    Gi = chol2inv(chol(G[b0,b0]))
    H[h1,h1] = G[b1,b1] - 
               t(G[b0,b1]) %*% Gi %*% G[b0,b1] + 
               t(G[b0,b1]) %*% Gi %*% H[h0,h0] %*% Gi %*% G[b0,b1]
    H[h1,h0] = t(G[b0,b1]) %*% Gi %*% H[h0,h0]
    H[h0,h1] = t(H[h1,h0])
    n0 = n1
  }
  return(H)
}

#-----------------------
doHinv = function(Glist) {
#-----------------------
  # return H inverse using algorithm in Legarra et al.
  # input is list of G matrices in increasing marker density
  nb = length(Glist)
  n  = ncol(Glist[[1]]) # total nind
  n1 = ncol(Glist[[2]])  # nind of lowest density array
  b0 = (n-n1+1):n  # inds genotyped 
  
  A = Glist[[1]] # NRM
  
  # Hinv is initialized to NRM inverse (brute force simply for demonstration purposes)
  Hinv = chol2inv(chol(A))
  
  # Only H of matrices except last one is needed
  H = doHgss(Glist[-1])
  H = chol2inv(chol(H)) - chol2inv(chol(A[b0,b0]))
  Hinv[b0,b0] = Hinv[b0,b0] + H
  return(Hinv)
}
```

## **Tutorial**

### **Define parameters**
```{r step00, exercise=TRUE}
# Parameters
h2 = 0.25       # heritability
nFounder = 100  # n founder individuals
nOff = 2        # n offspring / cross
nChr = 10       # n chromosomes
nQTL = 5        # nQTL / chr
nSNP = 1000     # num SNP / chr
nGen = 3        # no. of generations
```

### **Run simulation**
```{r step0, exercise=TRUE, exercise.setup="step00"}

# Simulation wrapper
sim <- runAlphasim(h2, nFounder=nFounder, nOffspring=nOff, nChr = nChr, nQTL=nQTL, nSNP = nSNP, nGen = nGen)

# saves sim output so can be read
saveRDS(sim, here("alphasim.rds"))
```

**By saving and re reading, same simulation data is used across code chunks**

```{r step1, exercise=TRUE, exercise.setup="step00"}
sim <- readRDS(here('alphasim.rds'))

# Returns 
g <- sim$g     # sequence data (whole population)
q <- sim$q
y <- sim$y     # phenotype
tbv <- sim$tbv # true breeding value
ped <- sim$ped # pedigree

nmkr <- ncol(g)
nind <- nrow(g)

print(paste0('N inds: ', nind, '; N mkrs: ', nmkr))
print(tail(y))
```

**Basic plots**
```{r step2, exercise=TRUE, exercise.setup="step1"}
hist(maf(g), main = 'Allele frequency distribution')

heatmap(cor(g[,1:100]), Rowv=NA, Colv=NA, main = 'LD map, first 100 markers')

plot(tbv, y, xlab='TBV', ylab='y')

```

### **Genotyping and individual blocks defined** 

They are specified as lists of markers and of markers. In this case, the first block contains  all individuals ungenotyped; the second block contains individuals from second generation genotyped for 20% markers, and a third block from last generation with all markers. Any number of blocks can be added.

**WARNING:** ids must be in same order in each block and $G$

```{r step3, exercise=TRUE, exercise.setup="step2"}
#--> blocks defined
b_ids = list(1:nind,                          # all individuals are in first block
             (nFounder*nOff+nFounder+1):nind, # array genotyping starts at third generation inds
             (nind-nFounder*nOff+1):nind)     # last generation is sequenced

#--> chip markers: contains marker ids included in each block array, 0 for pedigree
mk_ids = list(0,  # - first block has no markers, only pedigree
              sample(1:nmkr, nmkr%/%10), # - second block ids genotyped with random 10% markers
              1:nmkr) # - third block genotyped with all markers (sequence)


# no. of blocks
nblock = length(b_ids)

```

**Relationship matrices**

```{r step4, exercise=TRUE, exercise.setup="step3"}
#--> obtain relationship matrices
Glist = list()
for (b in 1:nblock) {
  if (length(mk_ids[[b]])==1) {
    # numerator relationship matrix
    Glist[[b]] = doA(ped)
  } else {
    # marker GRM for given inds and mkrs
    Glist[[b]] = GRM(g[b_ids[[b]], mk_ids[[b]]])
    diag(Glist[[b]]) = diag(Glist[[b]]) + 0.5 # to stabilize
  }
}
```

#### **GSS function**
**This is the function that obtains $H_{GSS}$**

```
#------------------------
doHgss <- function(Glist) {
#------------------------
  # Input is a list of molecular relationship matrices in increasing marker density
  nb = length(Glist)     # n G's
  n  = ncol(Glist[[1]])  # ninds
  H  = matrix(0, nrow=n, ncol=n)
  
  n0 = ncol(Glist[[nb]])
  H[(n-n0+1):n, (n-n0+1):n] = Glist[[nb]]   # inds for highest density genotyped
  for (i in (nb-1):1) {
    G  = Glist[[i]]   # G of current block
    n1 = ncol(G)      # coordinates wrt to current G
    b1 = 1:(n1-n0)    # individuals with current array ungenotyped with former denser array
    b0 = (n1-n0+1):n1 # previous individuals genotyped with denser array
    h1 = b1 + (n-n1)  # coordinates in global H: 1...nind
    h0 = b0 + (n-n1)
    Gi = chol2inv(chol(G[b0,b0]))
    H[h1,h1] = G[b1,b1] - 
               t(G[b0,b1]) %*% Gi %*% G[b0,b1] + 
               t(G[b0,b1]) %*% Gi %*% H[h0,h0] %*% Gi %*% G[b0,b1]
    H[h1,h0] = t(G[b0,b1]) %*% Gi %*% H[h0,h0]
    H[h0,h1] = t(H[h1,h0])
    n0 = n1
  }
  return(H)
}
```

```{r step5, exercise=TRUE, exercise.setup="step4"}
# Computes GSS H using defined Glist
H    <- doHgss(Glist)

# and its inverse
Hinv <- doHinv(Glist)

# Proof H * Hinv = I
I = H %*% Hinv
print(head(diag(I)))
print(round(I[1:10,1:10],10))

```

### **Individuals to be predicted are defined**

```{r step6, exercise=TRUE, exercise.setup="step5"}
# ids to be predicted are those with highest marker density
tst <- b_ids[[nblock]] 

```

### **Prediction with pedigree only**
```{r step7, exercise=TRUE, exercise.setup="step6"}
A = doA(ped)
blup_a = BLUP(A, y, tst, h2)
yHat = blup_a$uhat[tst]
c = cor(yHat, y[tst])
plot(y[tst], yHat, col='maroon3', main=paste0('Pedigree only, cor = ',round(c,3)))

```

### **Prediction with GSS**
```{r step8, exercise=TRUE, exercise.setup="step7"}
Hgss <- doHgss(Glist)
blup_gss <- BLUP(Hgss, y, tst, h2)
yHat <- blup_gss$uhat[tst]
c = cor(y[tst], yHat)
plot(y[tst], yHat, col='steelblue', main=paste0('GSS, cor = ',round(c,3)))
```

### **Prediction with SS**
Base population ungenotyped, rest genotyped with lower density array. Note the same function can be used, simply removing last block.

```{r step9, exercise=TRUE, exercise.setup="step8"}
Hgss <- doHgss(Glist[-nblock]) # last block removed
blup_ss <- BLUP(Hgss, y, tst, h2)
yHat <- blup_ss$uhat[tst]
c = cor(y[tst], yHat)
plot(y[tst], yHat, col='steelblue1', main=paste0('SS, cor = ',round(c,3)))
```

### **Prediction with complete sequence**
```{r step10, exercise=TRUE, exercise.setup="step9"}
G <- GRM(g)
blup_s <- BLUP(G, y, tst, h2)
yHat <- blup_s$uhat[tst]
c = cor(y[tst], yHat)
plot(y[tst], yHat, col='darkblue', main=paste0('Sequence, cor = ',round(c,3)))
```

**Write a code of your own**

```{r step11, exercise=TRUE, exercise.setup="step10"}


```

